---
title: "Concept attribution with a sparse autoencoder"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Concept attribution with a sparse autoencoder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

> **Status: experimental.** The concept layer described here ships no trained
> weights yet. Everything below runs once `inst/extdata/sae_weights_v1.rds`
> and `concept_labels_v1.csv` exist (see `dev/` for the training pipeline).

## What a sparse autoencoder is

An Alpha Earth embedding is a 64-dimensional vector that summarises everything
the satellites saw at a pixel over a year. Its dimensions are not individually
meaningful: the information about, say, canopy or moisture is smeared across
all 64 numbers. A sparse autoencoder (SAE) is a small network trained to
re-express those vectors in a much larger dictionary of directions — a few
hundred to a few thousand — under the constraint that only a handful (`k`) may
be active at any one pixel. Trained on millions of pixels sampled across every
continent and land-cover class, the directions that survive tend to align with
recurring kinds of land surface, because those are the patterns that make
reconstruction cheap.

Each dictionary entry is a *concept*: a direction in embedding space with an
activation you can map, and a label attached by correlating its activations
against catalog layers (climate, canopy height, land cover, terrain,
disturbance, night lights, phenology). The encoder runs entirely server-side
on Earth Engine as image math; nothing is trained or downloaded by you.

## What the concepts are — and are not

Concepts are **statistically labeled mixtures**, not measured variables. A
concept labeled "wetland, high soil moisture + low slope" is a direction whose
activation co-varies with those layers, not a wetland map. The label is a
convenience; the correlate table behind it (shipped in the label CSV) is the
ground truth, and `plot_concept_chips()` shows the top-activating locations so
you can check a label against actual imagery before leaning on it.

One caveat worth stating prominently: Alpha Earth has only 64 dimensions and
shows dimension-level redundancy, so SAE features may substantially overlap
with what per-dimension analysis already reveals. The value here is labels and
species-level attribution, not a claim that superposition is rampant in the
embeddings. The training pipeline therefore scores the SAE against NMF and
k-means dictionaries on the same sample — if the SAE does not beat those
baselines on label coherence, the shipped dictionary will be the baseline, and
this vignette will say so.

| dictionary | mean label coherence | dead latents |
|---|---|---|
| SAE (shipped config) | *pending Phase 4* | *pending* |
| NMF | *pending* | — |
| k-means | *pending* | — |

## Worked example: Bicknell's Thrush

```{r}
library(AlphaSDM)
setup_gee()

bith <- format_data(read.csv("bith_breeding.csv"),
                    coords = c("lon", "lat"), year = "year")

fit <- evaluate_models(bith, cv_folds = 5)

concepts <- derive_concepts(fit)
concepts
```

The expected top concepts for a Bicknell's Thrush breeding model look like
high-elevation dense conifer, krummholz, and cool-wet montane forest. If the
top concept is something incoherent, treat that as diagnostic signal about the
SAE, the labeling, or the SDM itself — it is worth chasing, not explaining
away.

Three attribution methods are available:

```{r}
# Selection (default): mean concept activation at presences over background,
# with a bootstrap CI. Works for every model.
derive_concepts(fit, method = "selection")

# Projection: cosine between a linear model's direction and each concept.
# Local and instant; only `similarity` models qualify.
fit_sim <- evaluate_models(bith, methods = "similarity", cv_folds = 0,
                           predict_coords = bith)
derive_concepts(fit_sim, method = "projection")

# Ablation: remove one concept from the embedding image, re-run the model,
# measure the mean suitability drop. The strongest causal reading, and the
# most expensive: each concept costs a full prediction pass, so it runs only
# for the top_n concepts ranked by selection.
derive_concepts(fit, method = c("selection", "ablation"), top_n = 10)
```

Verify a label before interpreting it:

```{r}
plot_concept_chips("C017")

maps <- attr(derive_concepts(fit, maps = TRUE), "maps")
# maps$activations is a lazy ee.Image; add to an rgee Map or export as usual.
```

## Temporal stability

The embeddings are annual. A concept that flips wildly between years over
stable land cover is suspect; check before trusting a multi-year analysis:

```{r}
a23 <- derive_concepts(fit, year = 2023)$selection_ratio
a22 <- derive_concepts(fit, year = 2022)$selection_ratio
cor(a23, a22, method = "spearman", use = "complete.obs")
```

## Cost notes

- `derive_concepts(method = "projection")`: no Earth Engine calls at all.
- `method = "selection"`: one activation-sampling pass at your presence points
  plus one background draw — comparable to scoring the same points during
  evaluation. Refused reads re-route through a batch export automatically.
- `method = "ablation"`: one full model-prediction pass over the region *per
  ablated concept*, reduced server-side. This is the expensive one; keep
  `top_n` modest and expect a batch-export wait on large regions.
- `maps = TRUE` attaches lazy `ee.Image` objects; nothing is computed until
  you display or export them. Displaying the full activation image at
  interactive zoom is heavy — select the concept bands you need.
