---
title: "Tuning and cross-validation with AlphaSDM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning and cross-validation with AlphaSDM}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>")
```

AlphaSDM is deliberately small: it fits a fixed **ensemble of complementary models**
on Alpha Earth embeddings, with all heavy computation on Google Earth Engine. It ships
sensible defaults rather than a built-in tuner, on the view that per-species tuning is
usually not where the gains are: the embeddings already encode most of the environmental
signal, and with few presences a cross-validated score is noisy enough that picking the
best configuration risks fitting that noise.

If you *do* want to tune hyperparameters or run cross-validation for your own data, you do
not need a special function; the package's regular building blocks are enough. This
vignette shows the common patterns. The whole public surface is short:

| function | purpose |
|---|---|
| `setup_gee()` | authenticate / initialise Earth Engine |
| `format_data()` | clean & standardise occurrence records |
| `evaluate_models()` | fit the model ensemble (optionally cross-validate / score a test set) |
| `generate_map()` | export a suitability raster |
| `calculate_classifier_metrics()`, `calculate_cbi()` | evaluation metrics (AUC, TSS, CBI, ...) |

## 1. The default ensemble

```{r}
library(AlphaSDM)
setup_gee(project = "your-gee-project")

occ <- format_data(my_records, coords = c("longitude", "latitude"),
                   year = "year", presence = "present")

# Fit the default curated ensemble and score an independent test set.
fit <- evaluate_models(data = occ, predict_coords = test_points,
                       aoi_year = 2024, cv_folds = 0)
fit$metrics                     # per-model + ensemble metrics on the test set
fit$metrics$ensemble$auc_roc    # e.g. the ensemble's AUC
```

## 2. Cross-validation

`evaluate_models()` has a built-in spatial cross-validation path. Set `cv_folds > 0` and
choose how folds are assigned with `cv_method`:

- `"block"`: spatial blocks (blockCV). Best estimate of **transfer** to new regions.
- `"kmeans"`: k-means clusters on coordinates (a lighter spatial split).
- `"random"`: non-spatial k-fold. Appropriate for **interpolation** within a sampled region.

```{r}
cv <- evaluate_models(data = occ, aoi_year = 2024,
                      cv_folds = 5, cv_method = "block")
cv$cv_metrics         # named list: cross-validated AUC per model (e.g. cv$cv_metrics$rf)
```

Pick `cv_method` to match your goal: spatial blocks if you will predict into unsampled
areas, random folds if your prediction targets are interspersed with your training data.

## 3. Hyperparameter tuning (rolling your own sweep)

Every model knob is a plain argument to `evaluate_models()` (e.g. `svm_cost`, `svm_gamma`,
`n_trees`, `shrinkage`, `max_nodes`, `maxent_beta`, `maxent_features`). To tune, loop over a
grid, cross-validate each setting, and keep the best by whatever metric you care about. This
reproduces what a built-in tuner would do, but transparently and only when you want it.

```{r}
grid <- expand.grid(svm_cost = c(10, 100), svm_gamma = c(0.05, 0.5))

scores <- apply(grid, 1, function(g) {
  res <- evaluate_models(data = occ, aoi_year = 2024, methods = "svm",
                         svm_cost = g[["svm_cost"]], svm_gamma = g[["svm_gamma"]],
                         cv_folds = 5, cv_method = "block")
  res$cv_metrics[["svm"]]      # cross-validated AUC for this configuration
})

grid[which.max(scores), ]   # best configuration by spatial-CV AUC
```

The same pattern tunes any method: swap in `n_trees`/`max_nodes`/`shrinkage` for `gbt`,
`maxent_beta`/`maxent_features` for `maxent`, and so on. To select among *methods* rather
than hyperparameters, vary the `methods` argument instead.

## 4. Custom ensembles

The default ensemble is `c("svm", "rf", "gbt", "maxent")`, a strong, complementary set.
You can fit any subset (or add the lighter models `similarity`, `knn`, `cart`, `mindist`)
by passing `methods`:

```{r}
# A leaner two-model ensemble
evaluate_models(data = occ, predict_coords = test_points, aoi_year = 2024,
                methods = c("rf", "maxent"), cv_folds = 0)
```

The recommended workflow is to start from the defaults and reach for the sections above
when you have a specific reason: an unusual dataset, a prediction target that differs from
the training region, or a need to report cross-validated performance.
